---
title: "200924_RWP26_L1_Intestine_FACS_DESeq2"
author: "RTPW"
date: "9/24/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Goal: Perform differential expression analysis on JM149 emrbyo intestine (GFP+) and non-intestine (GFP-) FACS sorted cells.

Plots to make:

- Correlation matrix of the input samples
- MA plot of 
  - whole embryo VS embryo cells
  - JM149 GFP+ cells vs JM149 GFP- cells

Run this first to install necessary packages

```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install()
# BiocManager::install("biomaRt")
# install.packages("tidyverse")
# install.packages("readxl")
# BiocManager::install("ComplexHeatmap")
# install.packages("matrixStats")
# install.packages("pheatmap")
# install.packages("RVAideMemoire")
# install.packages("dendextend")
# install.packages("binom")
# BiocManager::install("DESeq2")
# install.packages("corrplot")
# BiocManager::install("apeglm")
# install.packages("ashr")
```

Load necesseary packages

```{r}
# NOTE: BIOMART MUST BE LOADED BEFORE DPLYR
library(biomaRt)
library(DESeq2)
library(corrplot)
library(RColorBrewer)
library(pheatmap)
# library(dplyr)
# library(ggplot2)
# library(tibble)
library(tidyverse)
library(apeglm)
```
# Data input and processing

Read in the counts data

```{r}
countsData <- read.delim(file = "./all.counts", sep = " ")
head(countsData)
```
```{r}
colnames(countsData[6:15])
```


Read in and print metadata file
```{r}
metadata1 <- read.table(file = "./RWP27_metadata.tsv", header = FALSE, stringsAsFactors = FALSE)
colnames(metadata1) <- c("Filename.Fwd", "Filename.Rev", "names")

rep <- c(1,1,1,2,2,2,2,3,3,3)
type <- c("cells", "gut", "gutless", "whole", "cells", "gut", "gutless", "whole", "gut", "gutless")
metadata1 <- cbind(metadata1, rep, type)
metadata1
```

Factorize metadata
```{r}

metadata1$names <- factor(
  metadata1$names,
  levels = metadata1$names
)
metadata1$type <-
  factor(metadata1$type, levels = c("gutless", "gut", "cells", "whole"))
```

Order columns according to metadata1 order
```{r}
countsData <- countsData  %>% select(chr:length, sort(metadata1$names))
head(countsData)
```

Generate a table called "cts" out of the countsData table. Subset the countsData.
```{r}
cts <- as.matrix(countsData %>% select(metadata1$names))
head(cts)
```

Reorganize the metadata table so the names2 column are now headers

```{r}
rownames(metadata1)<- metadata1$names
coldata <- metadata1[,c("names", "rep", "type")]
rownames(coldata) <- as.vector(metadata1$names)
coldata
```

Check that the names match  --> Should be TRUE

```{r}
all(rownames(coldata) == colnames(cts))
```

# Contamination thresholding

Determine filtering threshold, identify read count threshold for non-intestine specific genes

```{r}
annotated_intestine_genes <- read.csv(file = "./C_elegans_All_Annotated_Intestine_Genes.tsv", header = TRUE, sep = "\t")
annotated_intestine_genes_smaller <- all_intestine_genes %>% dplyr::select(Gene.primaryIdentifier, Gene.symbol, Gene.expressionPatterns.lifeStages.publicName)
head(annotated_intestine_genes_smaller)
```


```{r}
tissue_specific_markers <- read.csv(file = "./Tissue_Specific_Marker_Genes.tsv", header = TRUE, sep = "\t")

tissue_specific_markers %>% 
  filter(ExpressionPattern.anatomyTerms.name %in% c("body wall musculature", "epithelial cell", "excretory cell", "intestine")) %>%
  dplyr::select(ExpressionPattern.genes.symbol, 
                ExpressionPattern.genes.primaryIdentifier, 
                ExpressionPattern.anatomyTerms.name)


tissue_specific_markers$gut[tissue_specific_markers$ExpressionPattern.anatomyTerms.name != "intestine"] = FALSE
tissue_specific_markers$gut[tissue_specific_markers$ExpressionPattern.anatomyTerms.name == "intestine"] = TRUE

tissue_specific_markers_smaller <- tissue_specific_markers %>%
  dplyr::select(ExpressionPattern.genes.symbol, 
                ExpressionPattern.genes.primaryIdentifier, 
                ExpressionPattern.anatomyTerms.name, gut)

cts_long <- as.data.frame(cts) %>% rownames_to_column(var = "WBGeneID") %>% 
  tidyr::pivot_longer(cols = embryoCells_rep1:embryoGFPminus_rep3, values_to = "reads") %>%
  tidyr::separate(name, sep = "_", into = c("sample_type", "rep")) 

cts_long_summary <- cts_long %>% 
  group_by(sample_type, WBGeneID) %>% 
  summarise(mean = mean(reads), variance = var(reads))
```


```{r}
cts_long %>% ggplot(aes(x = log(mean), y = log(variance))) +
  geom_point() +
  facet_wrap(~sample_type)
```

```{r}
cts_long %>% right_join(tissue_specific_markers_smaller, by = c("WBGeneID" = "ExpressionPattern.genes.primaryIdentifier")) %>% 
  ggplot(aes(x = log(mean), y = log(variance), color = gut)) +
  geom_point() +
  facet_wrap(~sample_type)
```
```{r}
cts_long %>% inner_join(annotated_intestine_genes_smaller %>% select(-Gene.expressionPatterns.lifeStages.publicName) %>%  unique(), by = c("WBGeneID" = "Gene.primaryIdentifier")) %>% 
  ggplot(aes(x = log(mean), y = log(variance))) +
  geom_point() +
  facet_wrap(~sample_type)


```

# Make DESeqDataSet

Generate the DESeqDataSet. The variables in this design formula will be the type of sample, and the preparation date. This should reduce the variability between the samples based on when they were made.

From the vignette: "In order to benefit from the default settings of the package, you should put the variable of interest at the end of the formula and make sure the control level is the first level."

The variable of interest is the sample `type`.

Using `DESeqDataSetFromMatrix` since I used the program `featureCounts`.

```{r}
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ type)
# filter lowly expressed genes
keep <- rowSums(counts(dds)) >=10
dds <- dds[keep,]
dds
```

# Perform Differential Expression

```{r}
dds <- DESeq(dds)
resultsNames(dds)
```

```{r}
res <- results(dds, contrast = c("type", "gut", "gutless"))
head(res)
```

Write results output file.  

```{r}
res_df <- as.data.frame(res)
# write.csv(x = res_df, "./200511_L1_intestine_FACS_gut_vs_gutless.csv", quote = FALSE)
```


```{r}
ma_gut_vs_gutless <- plotMA(res, ylim = c(-11,11), alpha = 0.05)

# pdf(file = "./EmbryoFACS_MA_plot_gut_vs_gutless_200930.pdf", 5, 5)
# plotMA(res, ylim = c(-11,11), alpha = 0.05)
# dev.off()
```

```{r}
resLFC <- lfcShrink(dds, coef = "type_gut_vs_gutless")
resApeglm <- lfcShrink(dds, coef = "type_gut_vs_gutless", type = "apeglm")
resAsh <- lfcShrink(dds, coef = "type_gut_vs_gutless", type = "ashr")
```


```{r}
par(mfrow = c(1,3), mar = c(4,4,2,1))
plotMA(resLFC, ylim=c(-10,10), main = "normal")
plotMA(resApeglm, ylim=c(-10,10), main = "apeglm")
# plotMA(resAsh, ylim=c(-10,10), main = "ashr")
```

```{r}
# write.csv(resApeglm %>% as.data.frame() %>% rownames_to_column(var = "WBGeneID"), file = "./200531_res_gut_vs_gutless_apeglmShrink.csv")

```

Export the plot

```{r}
# pdf(file = "./200707_L1FACS_Gut_vs_Nongut_MAplot.pdf", height = 5, width = 5)
# plotMA(resApeglm, ylim=c(-10,10), main = "L1 FACS Gut vs Nongut Differential Expression")
# dev.off()
```


# Sample-to-sample distance matrix

```{r}
vsd <- vst(dds, blind = FALSE)
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- vsd$names
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```

Export the plot

```{r}
# pdf(file = "./EmbryoFACS_Sample_Distance_Matrix_200930.pdf", height = 4, width = 6)
# pheatmap(sampleDistMatrix,
#          clustering_distance_rows = sampleDists,
#          clustering_distance_cols = sampleDists,
#          col = colors)
# dev.off()
```


# Principle Component Analysis

```{r}
plotPCA(vsd, intgroup = c("names"))
```


```{r}
plotCounts(dds, "WBGene00001250", intgroup = "type")
```


```{r}
rlog_counts <- assay(rlog(dds))
rlog_counts_df <- as.data.frame.matrix(rlog_counts) %>% rownames_to_column(var = "WBGeneID")
# write.csv(rlog_counts_df, file = "./200607_RWC24_L1intestineFACS_rlogCounts.csv")
```


# Intestine RNA vs ELT-2 Binding

```{r}
resApeglm_df <- as.data.frame.matrix(resApeglm) %>% rownames_to_column(var = "WBGeneID")

ap = readRDS(
  "../kw-onish-elt-2/ELT2_binding/annotatedPeaks.rds"
)
ap_mean <- ap %>% mcols %>% data.frame %>% group_by(feature) %>% summarize(LEoccupancy =
                                                                  mean(as.numeric(LE_nonNormed)))

head(ap_mean)

resApeglm_ap_mean_join <- ap_mean %>% full_join(resApeglm_df, by = c("feature" = "WBGeneID"))

lm(log10(baseMean) ~ LEoccupancy, 
   data = resApeglm_ap_mean_join %>% filter(log2FoldChange > 2, padj < 0.01)) %>% summary()
```

```{r}
ggplot(resApeglm_ap_mean_join %>% filter(padj < 0.01, log2FoldChange > 2), 
       aes(x = baseMean, y = LEoccupancy, color = log2FoldChange)) +
  geom_point() +
  scale_x_log10() +
  stat_smooth(method = lm) +
  scale_colour_gradientn(colours = rainbow(5))
```


```{r}
names(rlog_counts_df_apMean_join)
```

```{r}
rlog_counts_df_apMean_join <- ap_mean %>% full_join(rlog_counts_df, by = c("feature" = "WBGeneID"))

ggplot(rlog_counts_df_apMean_join %>% mutate(meanGfpPlus = (embryoGFPplus_rep1 + embryoGFPplus_rep2)/2), 
       aes(x = meanGfpPlus, y = LEoccupancy)) +
  geom_point() +
  stat_smooth(method = lm)

```

# Visualize non-intestine gene expression

```{r}
rlog_counts_long <- rlog_counts_df %>%
  pivot_longer(cols = embryoCells_rep1:embryoGFPminus_rep3, values_to = "reads") %>%
  separate(name, sep = "_", into = c("sample_type", "rep"))
  
rlog_counts_long_summary <- rlog_counts_long %>% group_by(sample_type, WBGeneID) %>% 
  summarise(mean = mean(reads), variance = var(reads))


rlog_counts_long_summary %>% inner_join(tissue_specific_markers_smaller, by = c("WBGeneID" = "ExpressionPattern.genes.primaryIdentifier")) %>%
  ggplot(aes(x = mean, y = variance, color = gut)) +
  geom_point() +
  facet_wrap(~sample_type)
```
```{r}
annotated_intestine_genes_smaller %>%
  filter(Gene.expressionPatterns.lifeStages.publicName %in% c("embryo Ce", 
                                                              "elongating embryo Ce",
                                                              "fully-elongated embryo Ce",
                                                              "3-fold embryo Ce",
                                                              "comma embryo Ce",
                                                              ""
                                                              ))

embryo_intestine_genes <- annotated_intestine_genes_smaller %>% filter(grepl("embryo", Gene.expressionPatterns.lifeStages.publicName)) %>% filter(Gene.expressionPatterns.lifeStages.publicName != "postembryonic Ce") %>%
  select(-Gene.expressionPatterns.lifeStages.publicName) %>%  unique()

rlog_counts_long_summary %>% inner_join(embryo_intestine_genes, by = c("WBGeneID" = "Gene.primaryIdentifier")) %>%
  ggplot(aes(x = mean, y = variance)) +
  geom_point() +
  facet_wrap(~sample_type)
```
```{r}
as.data.frame(res) %>% rownames_to_column(var = "WBGeneID") %>% 
  mutate(embryo_gut = if_else(
    condition = WBGeneID %in% embryo_intestine_genes$Gene.primaryIdentifier,
    true = TRUE,
    false = FALSE
  )) %>%
  ggplot(aes(x = log(baseMean), y = log2FoldChange, color = padj < 0.05)) +
  geom_point() +
  facet_wrap(~embryo_gut)
```
# Tissue specific ground truth

```{r}
ground_truth <- readxl::read_xlsx(path = "./Tissue_Specific_Ground_Truth_RTPW.xlsx")

neuron_truth <- ground_truth %>% rowwise() %>% mutate(neuron_count = sum(c_across(ADA:VD_DD))) %>% select(WBGeneID, gene_name, neuron_count)
neuron_truth



ubiquitous_genes <- neuron_truth %>% filter(neuron_count == ground_truth %>% dplyr::select(ADA:VD_DD) %>% ncol()) 
ubiquitous_genes

neuron_truth %>% inner_join(tissue_specific_markers, by = c("WBGeneID" = "ExpressionPattern.genes.primaryIdentifier"))


germline_genes <- data.frame(WBGeneID = c("WBGene00001598", "WBGene00003993", "WBGene00003992", "WBGene00010492"), gene_name = c("glh-1", "pgl-2", "pgl-1", "meg-1"))

germline_genes$tissue <- "germline"

ground_truth_markers <- ground_truth %>% mutate(tissue= case_when(
  gene_name == "elt-2" ~ "intestine",
  WBGeneID %in% (neuron_truth %>% inner_join(tissue_specific_markers, by = c("WBGeneID" = "ExpressionPattern.genes.primaryIdentifier")))$WBGeneID ~ "neuron",
  Intestine == 1 & Hypodermis == 0 & Muscle == 0 & Germline == 0 & Pharynx == 0 & Neuron == 0 ~ "intestine",
  Intestine == 0 & Hypodermis == 1 & Muscle == 0 & Germline == 0 & Pharynx == 0 & Neuron == 0 ~ "hypodermis",
  Intestine == 0 & Hypodermis == 0 & Muscle == 1 & Germline == 0 & Pharynx == 0 & Neuron == 0 ~ "muscle",
  Intestine == 0 & Hypodermis == 0 & Muscle == 0 & Germline == 1 & Pharynx == 0 & Neuron == 0 ~ "germline",
  Intestine == 0 & Hypodermis == 0 & Muscle == 0 & Germline == 0 & Pharynx == 1 & Neuron == 0 ~ "pharynx"
  )) %>% 
  dplyr::select(!(ADA:VD_DD)) %>%
  dplyr::select(!(Intestine:Neuron)) %>%
  drop_na(tissue) %>% 
  bind_rows(germline_genes)

ground_truth_markers$tissue <- factor(ground_truth_markers$tissue, levels = c("intestine", "hypodermis", "germline", "pharynx", "muscle", "neuron"))

tissues <- c("intestine", "hypodermis", "germline", "pharynx", "muscle", "neuron")
ordered_genes <- c()
for(i in tissues){
  genes <- (ground_truth_markers %>% filter(tissue == i))$gene_name
  ordered_genes <- append(ordered_genes, genes)
}
ordered_genes
fct_count(ground_truth_markers$tissue)
```

```{r}
rlog_counts_long_summary %>% inner_join(ground_truth_markers, by = "WBGeneID") %>%
  ggplot(aes(x = mean, y = variance, color = tissue)) +
  geom_point() +
  facet_wrap(~sample_type)
```
```{r}
as.data.frame(res) %>% rownames_to_column(var = "WBGeneID") %>% 
  left_join(ground_truth_markers, by = "WBGeneID") %>%
  ggplot(aes(x = log(baseMean), y = log2FoldChange, color = padj < 0.05)) +
  geom_point() +
  geom_hline(yintercept = c(2,-2), color = "red") +
  facet_wrap(~tissue)
```
```{r}
cts_long %>% right_join(ground_truth_markers, by = "WBGeneID") %>% 
  ggplot(aes(x = gene_name, y = log(reads), color = tissue)) +
  geom_point() +
  coord_flip() +
  scale_x_discrete(limits = rev(ordered_genes)) +
  facet_wrap(~sample_type)
```
```{r}
rlog_counts_long %>% right_join(ground_truth_markers, by = "WBGeneID") %>%
  ggplot(aes(x = gene_name, y = reads, color = tissue)) +
  geom_point() +
  coord_flip() +
  scale_x_discrete(limits = rev(ordered_genes)) +
  facet_wrap(~sample_type)
```
```{r}
res_df %>% rownames_to_column(var = "WBGeneID") %>% 
  right_join(ground_truth_markers, by = "WBGeneID") %>%
  ggplot(aes(x = gene_name, y = log2FoldChange, color = tissue, size = baseMean)) +
  geom_point() +
  coord_flip() +
  scale_x_discrete(limits = rev(ordered_genes))
```
```{r}
hist(((res_df %>% filter(log2FoldChange > 2.5, padj < 0.05))$baseMean))
hist(log((res_df %>% filter(log2FoldChange > 2.5, padj < 0.05))$baseMean))
```
```{r}
res_df %>% rownames_to_column(var = "WBGeneID") %>% right_join(tissue_specific_markers_smaller, by = c("WBGeneID" = "ExpressionPattern.genes.primaryIdentifier")) %>%
  ggplot(aes(x = log(baseMean), y = log2FoldChange, color = gut)) +
  geom_point() +
  geom_hline(yintercept = c(5.5, -2.5))
```
```{r}


res_df %>% rownames_to_column(var = "WBGeneID") %>% right_join(embryo_intestine_genes, by = c("WBGeneID" = "Gene.primaryIdentifier")) %>%
  ggplot(aes(x = log(baseMean), y = log2FoldChange, color = padj < 0.05)) +
  geom_point() +
  geom_hline(yintercept = c(5.5, -2.5))
```
```{r}
res_df %>% rownames_to_column(var = "WBGeneID") %>% 
  filter(log2FoldChange >5.5, padj <0.05) %>% 
  filter(!(WBGeneID %in% embryo_intestine_genes$Gene.primaryIdentifier))
```

```{r}
embryo_intestine_data <- rlog_counts_df %>% inner_join(res_df %>% rownames_to_column(var = "WBGeneID"), by = "WBGeneID") %>%
  mutate(intestine_expression = case_when(
    log2FoldChange >= 2.5 & padj <= 0.01 ~ "enriched",
    log2FoldChange <= -2.5 & padj <= 0.01 ~ "depleted",
    log2FoldChange <= 2.5 & log2FoldChange >= -2.5 & padj >= 0.01 ~ "no_sig_diff",
    TRUE ~ "no_sig_diff"
  )) %>%
  drop_na(padj)

embryo_intestine_data %>%
  ggplot(aes(x = log10(baseMean), y = log2FoldChange, color = intestine_expression)) +
  geom_point()
```

```{r}
write_csv(embryo_intestine_data, file = "./220127_Embryo_Intestine_RNAseq_Data.csv")
```


```{r}
# <- ap_mean %>% right_join(embryo_intestine_data, by = c("feature" = "WBGeneID"))


embryo_intestine_data_long <- embryo_intestine_data %>% pivot_longer(cols = c(embryoGFPplus_rep1,embryoGFPplus_rep2,embryoGFPplus_rep3), values_to = "reads") %>%
  tidyr::separate(name, sep = "_", into = c("sample_type", "rep")) %>% 
  dplyr::select(WBGeneID, sample_type, rep, reads, intestine_expression, log2FoldChange)


# embryo_intestine_data_long_summary <- embryo_intestine_data_long %>% group_by(sample_type, WBGeneID) %>% 
#   summarise(mean = mean(reads), variance = var(reads))
  
embryo_RNA_vs_ChIP <- embryo_intestine_data_long %>% 
  group_by(WBGeneID) %>% 
  summarize(reads_rep_mean = mean(reads), reads_rep_variance = var(reads)) %>%
  ungroup() %>%
  inner_join(embryo_intestine_data_long %>% dplyr::select(WBGeneID, intestine_expression) %>% distinct(), by = "WBGeneID") %>%
  left_join(ap_mean, by = c("WBGeneID" = "feature")) %>% drop_na(LEoccupancy) 


embryo_RNA_vs_ChIP %>%
  ggplot(aes(x = reads_rep_mean, y = LEoccupancy, )) +
  geom_point() +
  stat_smooth(method = lm) +
  facet_wrap(~intestine_expression)

```

```{r}
ggplot(embryo_RNA_vs_ChIP, aes(x = reads_rep_mean, y = LEoccupancy)) +
  stat_binhex(color = "white") +
  theme_bw() +
  scale_fill_gradient(low = "white", high = "blue") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

```{r}
ggplot(embryo_RNA_vs_ChIP, aes(x = reads_rep_mean, y = LEoccupancy)) +
  stat_binhex(color = "grey") +
  stat_smooth(method = lm) +
  theme_bw() +
  scale_fill_gradient(low = "white", high = "blue") +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  facet_wrap(~intestine_expression)
```

